---
title: "Re-analysis Kurd data"
author: "Kaspar Bresser"
date: "06/10/2021"
output: 
#  github_document:
#    toc: TRUE
  html_document: 
    theme: simplex
    highlight: pygments
    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")

```


In the DivisionRecorder manuscript we performed a re-analysis of data obtained by Kurd et al. (Science Immunology, 2020) in order to test if 'replicative quiescent' Tcm from an external data-set would resemble the low-division Tcm in our data-set. 

To test this we assessed Tcm clusters from the Kurd data for enrichment/depletion of a 'replicative quiescence' associated gene signature, and then calculated correlations between the most/least quiescent from the Kurd data and all MetaCells from our data-set. 

Load in packages used in the analysis, and import the data. We used the metacell package to cluster the data obtained by Kurd et al. We'll use 

```{r loading}
library(here)
library(metacell)
library(Seurat)
library(rstatix)
library(viridis)
library(scales)
library(tidytext)
library(tidyverse)


# point the metacell package to the data directory
scdb_init(here("Data"), force_reinit=T)

# import metacell objects
mc <- scdb_mc("D90_MC_col")
MC.graph <- scdb_mc2d("D90_MC_col")

# import UMI-count table from the Kurd data-set
UMI.counts <- read_tsv(here("Data", "Kurd_D90_counts.tsv"))
```


# Identify subsets

We want to do a focused analysis on Tcm, so we'll roughly cluster the  memory T cell MetaCells from the Kurd data-set into Tem and Tcm. 

## Process data

To do this we'll get gene expression information from the UMI count table, by first converting to a seurat object so we can use this package for normalization. In the current table genes are stored as columns, the `CreateSeuratObject` function takes a table in which cell-codes are columns, and genes are rows.

```{r make_seurat}
UMI.counts %>% 
  column_to_rownames("cell_id") %>% 
  t() %>% 
  CreateSeuratObject() -> seurat.Kurd

seurat.Kurd
```

Normalize data using Seurats build-in method.

```{r normalize}
seurat.Kurd <- NormalizeData(seurat.Kurd, assay = "RNA", normalization.method = "CLR")
```

We'll focus our analysis on a set of manually curated genes that relate to immune function. Import the set.

```{r import_genes}
immune.genes <- read_rds(here("Data", "imm_genes.rds"))

head(immune.genes, 20)
```

Next, extract the normalized UMI counts for these genes from the seurat object. 

```{r get_expression}
seurat.Kurd %>% 
  GetAssayData( slot = "data", assay = "RNA")%>% 
  as.matrix() %>%
  t() %>% 
  as_tibble(rownames = "cellcode") %>% 
  select(any_of(c("cellcode", immune.genes)))-> expression.table
```


Get coordinates from the graph object and join with the expression data.

```{r get_coords}
coords <- tibble( cellcode = names(MC.graph@sc_y),
                  sc.x = MC.graph@sc_x,
                  sc.y = MC.graph@sc_y)
coords

plot.data <- inner_join(coords, expression.table, by = "cellcode")
```

We'll look at a small set of core effector and multipotency genes, so lets define those. 

```{r set_genesets}
eff.genes <-c("Tbx21", "Prdm1", "Id2", "Gzmb", "Klrg1", "Cx3cr1", "Prf1", "Lgals1","Ifng")
mult.genes <- c("Eomes", "Il7r","Sell", "Tcf7", "Bcl2", "Ccr7","Myb", "Bach2")
```

Finally we can add the summed effector and multipotency signature scores to the table, did this by first scaling the genes within the signature, and then summing those values, also capped at a certain maximum to reduce the effect of outliers in visualization.


```{r calc_scores}
plot.data %>% 
  mutate(effector = rowSums( scale(select(., one_of(eff.genes))) ),
         effector.capped = case_when(effector > 10 ~ 10,
                                     TRUE ~ effector)) %>% 
  mutate(multipo = rowSums( scale(select(., one_of(mult.genes))) ),
         multipo.capped = case_when(multipo > 10 ~ 10,
                                     TRUE ~ multipo)) -> plot.data
```



## Visualize subsets

Lets first look at the MetaCell projection. Join the data with the MetaCell identities and plot. Take the MetaCell identities from the `mc` object. 

```{r plot_MC, fig.asp=.8}
enframe(as.factor(mc@mc), name = "cellcode", value = "MC") %>% 
  right_join(plot.data) %>% 
ggplot( aes(x=sc.x, y=sc.y, color = MC))+ 
  geom_point(size = 1)+
  theme_classic()+ 
  theme(legend.position = "right")+
  scale_color_viridis_d()
```


Next plot the effector signature


```{r plot_effector}
plot.data %>% 
  arrange(effector) %>% 
  ggplot( aes(x=sc.x, y=sc.y, color = effector.capped))+ 
  geom_point(size = 0.8)+
  scale_color_gradient2(low= "dodgerblue3", high= "red3", mid= "white", midpoint = 0)+
  theme_classic()


```

And same for the multipotency signature. 


```{r plot_multipotency}
plot.data %>% 
  arrange(multipo) %>% 
  ggplot( aes(x=sc.x, y=sc.y, color = multipo.capped))+ 
  geom_point(size = 0.8)+
  scale_color_gradient2(low= "dodgerblue3", high= "red3", mid= "white", midpoint = 0)+
  theme_classic()

```


Based on these data we defined MC5 as a Tem cluster, and the remaining MetaCells as Tcm clusters. 

Lets have a look at the expression of some genes in these newly defined subsets. 

```{r prep_violin}
plot.genes <- c("Bcl2", "Il7r", "Sell", "Klrg1", "Cx3cr1", "Gzmb")
plot.data %>% 
  left_join(enframe(as.factor(mc@mc), "cellcode", "MC")) %>% 
  mutate(phenotype = case_when(MC == "5" ~ "TEM",
                               TRUE ~ "TCM")) %>% 
  select(one_of(c(plot.genes, "MC", "phenotype")) ) %>% 
  pivot_longer(cols = -c("MC", "phenotype"),  names_to = "gene", values_to = "expression") %>% 
  mutate(gene = factor(gene, plot.genes))  -> for.violins

for.violins
```


```{r plot_violin}
ggplot(for.violins, aes(x = phenotype, y = expression, fill = phenotype))+
  geom_violin(scale = "width")+
  facet_wrap(~gene, scales = "free_y", nrow = 2)+
  scale_fill_manual(values = c("#0036A8","#C1272D"))
```

# Calculate QstemScore

Next we'll zoom in on the Tcm MetaCells of the Kurd data-set and calculate the relative expression of genes associated wiht replicative quiescence in stem cell, or QstemScore as we refer to it in the manuscript. 

The signature consists of genes negatively or positively associated with stem cell quiescence. Lets read in both.

```{r read_sigs}
gs.neg <- read_lines(here("Data", "Quiescence_neg_sig.txt"))
gs.pos <- read_lines(here("Data", "Quiescence_pos_sig.txt"))
```

We'll calculate a positive and a negative score, using the gene-enrichment values that are stored in the `mc` object.


```{r calc_Qstem1}
lfp <- log2(mc@mc_fp)


lfp %>% 
  as_tibble(rownames = "genes") %>% 
  filter(genes %in% c(gs.pos, gs.neg)) %>% 
  pivot_longer(cols = -c(genes), names_to = "MC", values_to = "lfp") %>% 
  mutate(direction = case_when(genes %in% gs.pos ~ "positive",
                               genes %in% gs.neg ~ "negative")) %>% 
  group_by(MC, direction) %>% 
  summarise(score = sum(lfp)) -> QstemScores
  

```

Then calculate a complete QstemScore by subtracting the negative score from the positive score

```{r calc_Qstem2}
QstemScores %>% 
  pivot_wider(names_from = "direction", values_from = "score") %>% 
  group_by(MC) %>% 
  summarise(score = positive - negative ) -> QstemScores

QstemScores
```


Now to visualize. Also, drop MC5 (the Tem MetaCell).

```{r plot_QstemScore, fig.asp=1, fig.width=3}
QstemScores %>% 
  filter(!(MC == 5)) %>% 
  mutate(MC = fct_reorder(MC, score)) %>% 
  ggplot(aes(MC, score))+
    geom_bar(stat = "identity", fill = "grey70", color = "black")+
    coord_flip()+
    theme_classic()+
    scale_y_continuous(breaks = breaks_width(0.5))+
    labs(x = "Tcm MetaCell Kurd data-set", y = "QstemScore")
```

# Calculate correlations

In the above analysis we found that MC3 and MC1 had the highest QstemScore, and MC6 and MC7 had the lowest. We'll check how well these MCs correlate, in terms of gene-expression of immune related genes, to the MCs found in our data-set. 

Import the mc object from the DivisionRecorder manuscript and extract the gene-enrichment values.

```{r get_DivRec_data}
mc.DivRec <- scdb_mc("DivRecMEM_MC")

lfp.DivRec <- log2(mc.DivRec@mc_fp)
```

We'll focus on immune related genes, that we have data on in both data-sets.

```{r focus_genes}
focus.genes <- intersect(row.names(lfp),row.names(lfp.DivRec))
focus.genes <- intersect(focus.genes, immune.genes)
```

Convert both tables to tibbles

```{r to_tibbles}
lfp <- as_tibble(lfp, rownames = "genes")
lfp.DivRec <- as_tibble(lfp.DivRec, rownames = "genes")
```

Join the tibbles, first select the MetaCells from the Kurd data that we're interested in, then join wiht the DivisionRecorder data. Also, rename both sets.

```{r combine_data, warning=F}
lfp %>% 
  select(one_of( c("genes", "1", "3", "6", "7"))) %>% 
  rename_with( ~paste0("Kurd", .), where(is_numeric) ) %>% 
  inner_join(lfp.DivRec) %>% 
  rename_with( ~paste0("DivRec", .), all_of(as.character(1:23)) ) -> combined.data

combined.data
```


Now calculate the correlations between Kurd MetaCells and the DivRec MetaCells, focusing on our selected genes. `rstatix::cor_test()` has a nice implementation to compare specific columns with each other using `tidyselect` verbs.

```{r correlations}
combined.data %>% 
  filter(genes %in% focus.genes) %>% 
  cor_test(vars = contains("Kurd"), vars2 =  contains("DivRec"), method = "pearson") %>% 
  adjust_pvalue(p.col = "p", output.col = "p.adj", method = "bonferroni") -> correlations

correlations
```


And finally plot the correlations. Lets highlight the significant correlations. 

```{r plot_corrs, fig.asp=1.1, fig.width=8}
correlations %>% 
  mutate(var2 = reorder_within(var2, cor, var1)) %>% 
  ggplot(aes(var2, cor, fill = p.adj < 0.05))+
    geom_bar(stat = "identity", color = "black")+
    facet_wrap(~var1, scales = "free")+
    coord_flip()+
    scale_x_reordered()+
    theme_classic()+
    labs(x = "DivisionRecorder manuscript MetaCells", y = "Pearson correlation")
```

And plot again, but this time color by multipotent or effector-like Tcm.

```{r plot_corrs2, fig.asp=1.1, fig.width=8}
Tcm.mult <- paste0("DivRec", c('9','1','12','13','14','16','11','2','3'))

correlations %>% 
  mutate(Tcm.subset = case_when(var2 %in% Tcm.mult ~ "mult",
                                TRUE ~ "eff"),
         var2 = reorder_within(var2, cor, var1)) %>% 
  ggplot(aes(var2, cor, fill =Tcm.subset))+
  geom_bar(stat = "identity", color = "black")+
  facet_wrap(~var1, scales = "free")+
  coord_flip()+
  scale_x_reordered()+
  scale_fill_manual(values = c("#4575B4" ,"#8AB8D7"))+
  theme_classic()+
  labs(x = "DivisionRecorder manuscript MetaCells", y = "Pearson correlation")
```





